### 2021-08-02 paper/localpsf.tex
edits

\section{Constructing the product-convolution H-matrix approximation}
\label{sec:method}
This section describes how we compute the convolution kernels, $\varphi_i$, and weighting functions, $w_i$, for the product-convolution approximation, $\AopPc \approx \Aop$, given in \eqref{eq:product_convolution}. We also describe how we convert $\AopPc$ to H-matrix format, and (optionally) how we modify $\AopPc$ to force it to be symmetric positive semi-definite.

To briefly summarize, we extract $\varphi_{x_i}$ for many points $x_i$ by applying $\Aop$ to Dirac combs associated with ``batches'' of points (Section \ref{sec:get_impulse_response}). 
%To reduce memory usage, all kernels $\varphi_i$ associated with a given batch, $b$, are stored in a single function, $\eta^b$, rather than as distinct functions. 
A greedy algorithm is used to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other, and are not too close to $\partial \Omega$ (Section \ref{sec:greedy_point_selection}). We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\phi_x$ (Section \ref{eq:mean_and_covariance_estimation}). The support of $\phi_x$ is estimated to be contained within an ellipsoid, so the process of choosing batches of points is an ellipse packing problem. For the weighting functions, we use the smoothest functions (in a least-squares sense) that interpolate the points $x_i$; constructing the weighting functions requires solving two Poisson PDEs per point $x_i$ (Section \ref{sec:weighting_functions}).
%\begin{itemize}
%\item We extract $\varphi_{x_i}$ for many points $x_i$ by applying $\mathcal{A}$ to Dirac combs associated with ``batches'' of points (Section \ref{sec:get_impulse_response}). To reduce memory usage, all kernels $\varphi_i$ associated with a given batch, $b$, are stored in a single function, $\eta^b$, rather than as distinct functions.
%\item We use a greedy algorithm to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other, and are not too close to $\partial \Omega$ (Section \ref{sec:greedy_point_selection}). 
%\item We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\psi_x$ (Section \ref{eq:mean_and_covariance_estimation}). 
%\item For the weighting functions, we use the smoothest functions (in a least-squares sense) that interpolate the points $x_i$; constructing the weighting functions requires solving two Poisson PDEs per point $x_i$ (Section \ref{sec:weighting_functions}).
%\end{itemize}
The complete algorithm for constructing $\AopPc$ is shown in Algorithm \ref{alg:construct_Atilde}. The method for evaluating kernel entries of $\Phi_\text{pc}$ is shown in Section \ref{sec:eval_matrix_entries}. The method for converting $\AopPc$ to H-matrix format is shown in Section \ref{sec:H_matrix_conversion}. The method for modifying the resulting H-matrix to make it symmetric positive semi-definite is shown in Section \ref{sec:make_spd}.

@@

\subsection{Sample point selection}

@@

The ellipsoid $E_x$ is not immediately available. To use $E_x$, we must first compute the scalar $\alpha(x)$, the length-$d$ vector $\mu(x)$, and the $d \times d$ matrix $\Sigma(x)$ (recall $d \in \{1,2,3\}$ is the spatial dimension of $\Omega$). The direct approach to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\phi_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\phi_x$ to determine $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$. But this direct approach is computationally infeasible because and it requires applying $\Aop$ to one vector per point $x$ that we wish to evaluate $E_x$ at, and we will need to evaluate $E_x$ at a large number of points. Fortunately, it is possible to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$, \emph{for all points $x$ simultaneously}, by applying $\Aop^T$ to one constant function, $d$ linear functions, and $d(d+1)/2$ quadratic functions. We present this efficient method of computing $\alpha$, $\mu$, and $\Sigma$ in Theorem \ref{thm:vol_mean_cov} and summarize the method in Algorithm \ref{alg:varhpi_mean_cov}.

@@

Now that we have estimates of the supports of the functions $\phi_x$, we use these estimates, combined with a greedy ellipsoid packing algorithm, to choose sets of sample points, $S_b$. These sample points will be used in Section \ref{sec:get_impulse_response} to form Dirac combs, which are then used to compute impulse responses of $\mathcal{A}$. We start with a finite set of candidate points $P$. To build a batch of sample points, $S_b$, first we initialize $S_b$ as an empty set. Then we select the point $p \in P$ that is the farthest away from all points in previous sample point batches. That is, $p$ is a maximizer of the following optimization problem:
\begin{equation*}
\max_{p \in P} \min_{x \in S_1 \cup \dots \cup S_{b-1}} \|p - x\|.
\end{equation*}
Once $p$ is selected, we remove $p$ from $P$. If $p$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_p \cap E_q = \{\}$ for all $q \in S_b$, and if $p$ is sufficiently far from the boundary, in the sense that $E_p \cap \partial \Omega = \{\}$, then we add $p$ to $S_b$. Otherwise we discard $p$. Recall that $E_p$ and $E_q$ are ellipsoids associated with points $p$ and $q$, respectively, as defined in \eqref{eq:support_ellipsoid}. This process repeats until there are no more points in $P$.  This is detailed in Algorithm \ref{alg:point_choice}. We determine whether $E_p \cap E_q = \{\}$ using the fast ellipsoid intersection test described in Appendix \ref{sec:fast_ellipsoid_intersection_test}.

@@

When $x_i$ is near a boundary, it may occur that some gridpoints for the regular grid are in locations where $\varphi_{x_i}$ is undefined. In this case, we store a placeholder value for $\varphi_{x_i}$ at those gridpoints. We use the IEEE standard value of ``NaN'' (not a number) for this placeholder. These NaN values are later replaced with more appropriate values, as we will discuss in Section \ref{sec:boundary_extension}.

@@

While $\varphi_{x_i}$ may be undefined at a point $z$, often there is a sample point $x_j$ that is near $x_i$, but further away from the boundary than $x_i$, such that $\varphi_{x_j}$ is well defined at $z$. To address boundary artifacts, we fill in missing/undefined portions of convolution kernels using information from other convolution kernels associated with the $k$ nearest sample points. Kernels associated with nearer sample points are given precedence over kernels associated with farther away sample points. 
%After the convolution kernel has been extended using information from neighboring convolution kernels, it is then further extended by zero.

In detail, let $p_1, p_2, \dots, p_k$ be the $k$ nearest sample points to $x_i$, ordered by nearness, including $x_i$ itself. That is, $p_1=x_i$, $p_2$ is the nearest neighbor to $x_i$, $p_3$ is the second nearest neighbor to $x_i$, and so on. Also, recall that in computations that the functions $\varphi_{x_i}$ are stored on regular rectilinear grids. To extend $\varphi_{x_i}$, first we form an expanded regular rectilinear grid that has the same grid spacing as the grid used for $\varphi_{x_i}$, and contains that grid as a subgrid, but is large enough to contain the boxes associated with all of the neighbor kernels, $\varphi_{p_1}, \varphi_{p_2}, \dots, \varphi_{p_k}$. We initialize an extended convolution kernel, $\varphi_{x_i}^E$, to zero on this expanded grid. Then we interpolate the functions $\varphi_{p_j}$ onto this expanded grid one at a time in reverse order. That is, first we interpolate $\varphi_{p_k}$ onto the grid, then $\varphi_{p_{k-1}}$, and so on until we reach the original kernel $\varphi_{p_1} = \varphi_{x_i}$. At each step, we overwrite existing values of $\varphi_{x_i}^E$ with new values from the kernel currently being interpolated, except when the new value of $\varphi_{p_j}$ is undefined (``NaN''). At gridpoints where the kernel being interpolated is undefined, we leave $\varphi_{x_i}^E$ unchanged. In this way, we fill in undefined values of $\varphi_{x_i}$ with values from successively further and further away kernels. This process is illustrated in Figure FIG. To avoid notational clutter, and since we will henceforth only work with the extended kernel, $\varphi_{x_i}^E$, in the remainder of this paper the symbol $\varphi_{x_i}$ refers to $\varphi_{x_i}^E$.

@@

@@

\subsection{Postprocessing the product-convolution approximation}

\subsubsection{Hierarchical matrix construction}
\label{sec:H_matrix_conversion}

@@

\subsubsection{Rational positive semi-definite modification}
\label{sec:make_spd}

>>>>

\section{Constructing the product-convolution H-matrix approximation}
\label{sec:method}
Here we describe how we choose the sample points, $x_i$ (Section \ref{sec:sample_point_selection}), how we form the convolution kernels, $\varphi_i$ (Section \ref{sec:get_impulse_response}), and how we form the weighting functions, $w_i$ (Section \ref{sec:weighting_functions}), for the product-convolution approximation, $\AopPc \approx \Aop$, given in \eqref{eq:product_convolution}. We also describe how we convert $\AopPc$ to H-matrix format (Section \ref{sec:H_matrix_conversion}), and (optionally) how we modify the resulting H-matrix, $\AmatPc$, to force it to be symmetric positive semi-definite (Section \ref{sec:make_spd}). The complete algorithm for constructing $\AmatPc$ is shown in Algorithm \ref{alg:construct_Atilde}. 
%The method for evaluating kernel entries of $\Phi_\text{pc}$ is shown in Section \ref{sec:eval_matrix_entries}.

%\begin{itemize}
%\item We extract $\varphi_{x_i}$ for many points $x_i$ by applying $\mathcal{A}$ to Dirac combs associated with ``batches'' of points (Section \ref{sec:get_impulse_response}). To reduce memory usage, all kernels $\varphi_i$ associated with a given batch, $b$, are stored in a single function, $\eta^b$, rather than as distinct functions.
%\item We use a greedy algorithm to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other, and are not too close to $\partial \Omega$ (Section \ref{sec:greedy_point_selection}). 
%\item We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\psi_x$ (Section \ref{eq:mean_and_covariance_estimation}). 
%\item For the weighting functions, we use the smoothest functions (in a least-squares sense) that interpolate the points $x_i$; constructing the weighting functions requires solving two Poisson PDEs per point $x_i$ (Section \ref{sec:weighting_functions}).
%\end{itemize}
%The complete algorithm for constructing $\AopPc$ is shown in Algorithm \ref{alg:construct_Atilde}. The method for evaluating kernel entries of $\Phi_\text{pc}$ is shown in Section \ref{sec:eval_matrix_entries}. The method for converting $\AopPc$ to H-matrix format is shown in Section \ref{sec:H_matrix_conversion}. The method for modifying the resulting H-matrix to make it symmetric positive semi-definite is shown in Section \ref{sec:make_spd}.

@@

\subsection{Sample point selection}
\label{sec:sample_point_selection}

We choose sample points, $x_i$, in batches. A greedy algorithm is used to choose as many points as possible per batch, while ensuring that the points in each batch are not too close to each other. We ensure that the points $x_i$ are well separated from each other by forming a-priori estimates of the supports of the functions $\phi_x$ (Section \ref{eq:mean_and_covariance_estimation}). The support of $\phi_x$ is estimated to be contained within an ellipsoid, so the process of choosing batches of points is an ellipsoid packing problem (Section \ref{sec:greedy_point_selection}). 
%In the next section (Section \ref{sec:get_impulse_response}), we will extract $\varphi_{x_i}$ for all the points $x_i$ in a given batch by applying $\Aop$ to the Dirac comb associated with that batch.

@@

The ellipsoid $E_x$ is not immediately available. To use $E_x$, we must first compute the scalar $\alpha(x)$, the length-$d$ vector $\mu(x)$, and the $d \times d$ matrix $\Sigma(x)$ (recall $d \in \{1,2,3\}$ is the spatial dimension of $\Omega$). The direct approach to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\phi_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\phi_x$ to determine $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$. But this direct approach is computationally infeasible because and it requires one operator action of $\Aop$ per point $x$ that we wish to evaluate $E_x$ at, and we will need to evaluate $E_x$ at a large number of points. Fortunately, it is possible to compute $\alpha(x)$, $\mu(x)$, and $\Sigma(x)$, \emph{for all points $x$ simultaneously}, by applying $\Aop^T$ to one constant function, $d$ linear functions, and $d(d+1)/2$ quadratic functions. We present this efficient method of computing $\alpha$, $\mu$, and $\Sigma$ in Theorem \ref{thm:vol_mean_cov} and summarize the method in Algorithm \ref{alg:varhpi_mean_cov}.

@@

Now that we have estimates of the supports of the functions $\phi_x$, we use these estimates, combined with a greedy ellipsoid packing algorithm, to choose sets of sample points, $S_b$. These sample points will be used in Section \ref{sec:get_impulse_response} to form Dirac combs, which are then used to compute impulse responses of $\mathcal{A}$. We start with a finite set of candidate points $P$. To build a batch of sample points, $S_b$, first we initialize $S_b$ as an empty set. Then we select the point $p \in P$ that is the farthest away from all points in previous sample point batches. That is, $p$ is a maximizer of the following optimization problem:
\begin{equation*}
\max_{p \in P} \min_{x \in S_1 \cup \dots \cup S_{b-1}} \|p - x\|.
\end{equation*}
Once $p$ is selected, we remove $p$ from $P$. If $p$ is sufficiently far from all of the previously chosen points in the current batch, in the sense that $E_p \cap E_q = \{\}$ for all $q \in S_b$,
%and if $p$ is sufficiently far from the boundary, in the sense that $E_p \cap \partial \Omega = \{\}$, 
then we add $p$ to $S_b$. Otherwise we discard $p$. Recall that $E_p$ and $E_q$ are ellipsoids associated with points $p$ and $q$, respectively, as defined in \eqref{eq:support_ellipsoid}. This process repeats until there are no more points in $P$.  This is detailed in Algorithm \ref{alg:point_choice}. We determine whether $E_p \cap E_q = \{\}$ using the fast ellipsoid intersection test described in Appendix \ref{sec:fast_ellipsoid_intersection_test}.

@@

When $x_i$ is near a boundary, it may occur that some gridpoints for the regular grid are in locations where $\varphi_{x_i}$ is undefined. In this case, we store a placeholder value for $\varphi_{x_i}$ at those gridpoints. We use the IEEE standard value of ``NaN'' (not a number) for this placeholder. Later we replace these NaN values with more appropriate values (Section \ref{sec:boundary_extension}).

@@

While $\varphi_{x_i}$ may be undefined at a point $z$, often there is a sample point $x_j$ that is near $x_i$, but further away from the boundary than $x_i$, such that $\varphi_{x_j}$ is well defined at $z$. To address boundary artifacts, we fill in missing/undefined portions of convolution kernels using information from other convolution kernels associated with the $k$ nearest sample points (we use $k=8$ in our numerical results). Kernels associated with nearer sample points are given precedence over kernels associated with farther away sample points. 
%After the convolution kernel has been extended using information from neighboring convolution kernels, it is then further extended by zero.

In detail, let $p_1, p_2, \dots, p_k$ be the $k$ nearest sample points to $x_i$, ordered by nearness, including $x_i$ itself. That is, $p_1=x_i$, $p_2$ is the nearest neighbor to $x_i$, $p_3$ is the second nearest neighbor to $x_i$, and so on. Also, recall that in computations that the functions $\varphi_{x_i}$ are stored on regular rectilinear grids. To extend $\varphi_{x_i}$, we form an expanded regular rectilinear grid that has the same grid spacing as the grid used for $\varphi_{x_i}$, and contains that grid as a subgrid, but is large enough to contain the boxes associated with all of the neighbor kernels, $\varphi_{p_1}, \varphi_{p_2}, \dots, \varphi_{p_k}$. We initialize an extended convolution kernel, $\varphi_{x_i}^E$, to zero on this expanded grid. Then we interpolate the functions $\varphi_{p_j}$ onto this expanded grid in reverse order. First we interpolate $\varphi_{p_k}$ onto the grid, then $\varphi_{p_{k-1}}$, and so on until the last step in which we interpolate the original kernel $\varphi_{p_1} = \varphi_{x_i}$. At each step, we overwrite existing values of $\varphi_{x_i}^E$ with new values from the kernel currently being interpolated, except when the new value of $\varphi_{p_j}$ is undefined (``NaN''). At gridpoints where the kernel being interpolated is undefined, we leave $\varphi_{x_i}^E$ unchanged. In this way, we fill in undefined values of $\varphi_{x_i}$ with values from successively further and further away kernels. This process is illustrated in Figure FIG. To avoid notational clutter, and since we will henceforth only work with the extended kernel, $\varphi_{x_i}^E$, in the remainder of this paper the symbol $\varphi_{x_i}$ refers to $\varphi_{x_i}^E$.

@@

For the weighting functions, we use the smoothest functions (in a least-squares sense) that interpolate the points $x_i$. Constructing the weighting functions requires solving two Poisson PDEs per sample point $x_i$.

@@

\subsection{Hierarchical matrix construction}
\label{sec:H_matrix_conversion}

@@

\subsection{Rational positive semi-definite modification}
\label{sec:make_spd}
